\documentclass[12pt,a4paper]{book}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage{url}
\usepackage{graphicx}

\usepackage[noend]{algorithmic}

% Laying out data structures.

\usepackage{bytefield}
% We use the bytefield package for describing data structures.  By
% default, it lays out fields from left to right.  We want the
% reverse.  It's easy enough to fix the bit header, however, when
% laying out a word, we have to put the first field in a word last.
%
% For example:
%
%   [  foo  | bar |    baz   ]
%
% is: foo % bar % baz
%
% even though baz starts at bit position 0.
% An environment for describing a struct.

% The default width per-bit is not large enough for a single letter.
% We want to be able to store a single capital letters.  This is good:
%
% \AtBeginDocument{\settowidth{\bitwidth}{W}}
%
% This maximizes the space according to the page width.
\AtBeginDocument{\setlength{\bitwidth}{0.98\textwidth / 32}}

\newcounter{structfoo}
\newenvironment{struct}[1]
        {\begin{center}\begin{bytefield}{#1}\setcounter{structfoo}{#1 - 1}\bitheader[b]{0,\value{structfoo}}\\}
        {\end{bytefield}\end{center}}

% Wrapper for the \bitbox macro that also displays the bit field's
% length.
\newcommand{\bits}[2]{\bitbox{#1}{\small{#2}\tiny{$_{(#1)}$}}}
% Name as above but the second argument is the label to use for
% displaying the bits.
\newcommand{\bitsl}[3]{\bitbox{#1}{\small{#3}\tiny{$_{(#2)}$}}}
% Wrapper for the \bitbox macro that also displays the bit field's
% length for 32-bit and 64-bit hosts.
\newcommand{\bitsx}[3]{\bitbox{#1}{\small{#3}\tiny{$_{(#1/#2)}$}}}
% Wrapper for the \bitbox macro that also displays the bit field's length.
\newcommand{\bit}[1]{\bitbox{1}{\small{#1}}}
\newcommand{\dontcare}[0]{$\sim$}


% Code snippets.

\usepackage{listings}
\lstloadlanguages{C}
\lstset{language={C}, columns=fullflexible}
\lstset{morekeywords={uint32_t, int32_t, uint16_t, int16_t,
    uintptr_t, intptr_t, bool, errot_t}}
\lstset{morekeywords={addr_t, addr_trans, object_policy, cap_properties}}

% Some convenience macros.
\newcommand{\keyword}[1]{\lstinline!#1!}
\newcommand{\func}[1]{\lstinline!#1!}
\newcommand{\var}[1]{\emph{#1}}
\newcommand{\type}[1]{\lstinline!#1!}
\newcommand{\const}[1]{\lstinline!#1!}
\newcommand{\errno}[1]{\lstinline!#1!}



\usepackage{makeidx}
\usepackage{tocbibind}
\usepackage[pagebackref,plainpages=false]{hyperref}

\title{Viengoos Developer Reference}
\author{Neal H. Walfield}
\date{January 2008}

\begin{document}

\frontmatter
\maketitle
\tableofcontents

\mainmatter

\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}

\chapter{Introduction}

The text you are reading describes the Viengoos virtual machine.  This
text is an attempt to provide a normative reference for Viengoos, to
enumerate its interfaces and to describe their behavior.  It also
attempts to explain the interfaces, to illustrate the motivation
behind some decisions and to show the interfaces' intended uses.  This
interleaving of the prescriptive with the descriptive may be a source
of confusion.  This is unintentional and as this document evolves,
such confusion will hopefully be eliminated.

\section{Overview}

Viengoos is an extensibility, object-capability system, ala Hydra
\cite{wulf74hydra} and EROS \cite{shapiro99eros}.  

Viengoos was built on the following ideas:

\begin{itemize}
\item object based,
\item virtualizable interfaces,
\item object statelessness \cite{tullmann96userlevel-checkpointing-through-exportable-kernel-state},
\item no kernel dynamic allocation,
\item resource accountability,
\item atomic methods,
\item caching \cite{cheriton94caching-model-of-os-kernel-functionality}
\item interrupt model \cite{ford99interface-and-execution-models},
\item activation-based \cite{roscoe95structure-of-a-multi-service-os}, and
\item resilience to destructive interference \cite{miller06robust-composition}
\end{itemize}

\subsection{Virtualizable Interfaces}

By virtualizable interfaces, we mean that all kernel implemented
objects can be easily proxied by a user-space task in such a way that
the proxy behaves in a manner indistinguishable from the kernel
implementation.

The idea is perhaps more easily explained through an example of a
familiar object that is not easily virtualizable.  Consider how files
are implemented on Unix-like systems and suppose that one process
wishes to proxy access to a file.  The proxy can open the file itself
and then provide another file descriptor to its clients.  The question
is what sort of file descriptor.  A pipe could be used.  In this case,
the proxy will see the clients' reads and writes, however, a pipe does
not support seeking and the kernel provides no way for the proxy to
interpose on this operation and provide its own implementation.

\subsection{Object Statelessness}

Continuing the previous example, supposing that there was a way to
cause such file invocations to be redirected to the proxy, another
problem arises: proxying a file is non-trivial as the object's state
machine is quite complicated.  For instance, the proxy must maintain a
file pointer for each client.  This is because each client expects
that the file descriptor it designates the normal file and that the
file pointer is private.  To work around this, the proxy must maintain
a private file pointer for each client and then serialize access to
the object and adjust the object's file pointer using the seek method
before invoking the read method.  This is required even if the proxy
only wants to do some sort of bounds checking.  To simplify
virtualization, objects should avoid maintaining sessions: as much as
is feasible, interfaces should be so designed such that a method
either senses or transforms the state of the object.

\section{Future Directions or TODO}

The objects and interfaces described in this document (mostly) reflect
the current implementation.  There are a number of limitations
requiring some thought.  The issues are outlined here.

\subsection{Virtualization}

Many of the methods are not virtualizable.  Indeed, two of them are
not even methods: cap\_copy and cap\_read are not invoked on an object
but are essentially system calls.  These should perhaps be modelled as
thread object methods.

However, there is a more complicated problem and that is: the kernel
walks the cappages and other objects to resolve an address but what
should it do when it encounters an end point?  The kernel cannot
invoke it as then it must wait for a reply and this provides an
opportunity for destructive interference.  The kernel also cannot
reply and to tell the client to revert to some other method of
resolution (or can it?).  If we just fault, a process can determine
whether an object is kernel or user-implemented by installing it in
its address space and then trying to access it.  Perhaps, the process
that does the virtualization can do some tricks to provide a
user-object but when the user tries to use it in its address space, by
interposing on the thread object, it can the install an appropriate
cappage.

As for the rest, to be able to virtualize a kernel object, the
implementation needs to have access to all the information that the
kernel implementation requires.  This means that we either have to
deliver more information or we have to adapt the interfaces.

An example of the former is to virtualize object\_slot\_copy\_in, the
implementation needs access to the source capability's subpage
descriptor, policy and the weak predicates.  This does not pose a
fundamental problem: making this information available to the object
does not violate POLA.  It does require that when a message is
delivered, any transferred capabilities must also include this
information.

Another example is invoking an object: the object implementation also
needs this information, in particular, the weak predicate and the
subpage descriptor. 

One place where the interfaces need to be change in a more fundamental
way is object\_slot\_copy\_out.  The kernel implementation of this
method does a bit copy from the designated source slot to the
designated target slot.  A user-implementation cannot do such a bit
copy; it needs to return the capability in the reply message.  To fix
this, the target parameter needs to be a return value and
object\_slot\_copy\_out simply returns an appropriate capability.
This raises another problem: how to then store the capability.  If we
store it in the receive buffer, then we still need another copy to get
it where we want.  If we store it directly in the address space (using
a scatter/gather technique), then we need to consider the case where
the target object is virtualized.

Also, for end points, we use the weak predicate to determine whether
the capability interface designates the send facet or the receive
facet.  To properly virtualize a kernel object, we need to allow the
user to control the weak predicate as normal.

\section{Outline}

This reference is divided into two parts.  The first describes the
kernel, how object addressing works, the primordial objects and
resource management mechanisms and policies.  The second part
describes the sample run-time environment shipped with Viengoos.

\include{viengoos}
\include{runtime}

% XXX: Without this \part, the bibliograph shows up as a chapter in
% the run-time \part.  With the \part, it at least shows up on the
% first level of the table of contents, although we then waste two
% pages and repeat \bibliography.
\part{Bibliograph}

\bibliographystyle{alpha}
\bibliography{bib}

\backmatter

% \printindex

\end{document}
